Adaptation Writ Large:\\ An Essay in Memory of John H. Holland

### 2015-10-13 paper.tex
Version with learning section plus other edits added

@@

>>>>

While genetic algorithms are nowadays often associated with models of
biological evolution, in Holland's view the processes of adaptation he
was studying manifested themselves in many different complex systems.
One system of special interest to Holland was the human mind, and in
particular, its capacity to learn via induction.

In the 1970s and 80s Holland formulated and extensively explored a
particular model of decision-making, action, and inductive learning,
called the {\emph classifier system}
(e.g., \cite{Holland&Reitman1978,EscapingBrittlenessPaper}.  In a
classifier system, a population of ``if-then'' rules interacts with an
environment, and over time is able to learn to improve its performance
via both {\emph credit-assignment} and {\emph rule-discovery}
algorithms.

Like the genetic algorithm, classifier systems can be viewed both as a
model of adaptation and as an artificial-intelligence (AI) method.  At
the time Holland was developing classifier systems, the field of AI
was focused on expert systems, which typically did not learn on their
own, a fundamental deficiency in Holland's view: ``[Expert] systems
are brittle in the sense that they respond appropriately only in
narrow domains, requiring substantial human intervention to compensate
for even slight shifts in domain'' \cite{EscapingBrittleness}. In
contrast, Holland proposed that ``{\emph induction} is the basic, and
perhaps only, way of making large advances in this direction.''

Holland's inspiration for classifier systems came from several
different disciplines.  Knowledge representation in the form of a
population of ``if-then'' rules seemed like a good choice, not only
because of its popularity in AI at the time but also from Holland's
early work on modeling Hebbian cell assemblies: ``In Hebbs view, a
cell assembly makes a simple statement: If such and such an event
occurs, then I will fire for a while at a high
rate.'' \cite{Quoted-in-Waldrop-Book-P.182}.

The if-then rules, when activated, compete to post their results on a
shared ``message list'', modeling the system's short-term memory
(again inspired by Hebb's work).  The competition was based on an
bidding process, in which a rule's current strength affected its
ability to bid highly.  Successful rules were gradually strengthened
(and unsuccessful rules weakened) via a credit-assignment method
called the {\emph bucket-brigade} algorithm, in which rules gaining
rewards from the environment or from other rules would pay back part
of their gains to those earlier-firing ``stage-setting'' rules that
set up the conditions for the eventual reward.  The idea for the
bucket-brigade algorithm was inspired by the mechanisms of Hebbian
learning and by early work on learning and optimization in economics,
psychology, control theory, and other fields (e.g.,
\cite{Bellman1957,Samuel1959}). \footnote{Of course similar
  credit-assignment methods have been used extensively in
  reinforcement learning, neural networks, and other areas of modern
  machine learning, but Holland was among the first to propose such
  methods.}

Finally, new rules for the population are discovered by a genetic
algorithm, using the strength of rules as a fitness function.

Classifier systems are fairly complex architectures, with learning
occurring at multiple time scales.  Nevertheless, such systems have
been usefully applied to areas as diverse as
poker-playing \cite{Smith1980}, control of gas pipeline
transmission \cite{Goldberg1983}, and modeling the stock
market \cite{PalmerEtAl}.  Modern reinforcement learning methods have
largely taken over as more useful approaches to real-world decision
and control problems, but classifier systems can perhaps be thought of
as an essential ``stage-setting'' method that enabled the development
of later approaches.

John Holland's main interest in classifier systems was how the two
learning mechanisms could work together to create subsets of rules
that accurately model the environment, and thus are able to ``fire''
at appropriate times.  An important part of such internal models is
the notion of a {\emph default hierarchy}---a set of rules that
together form a kind of hierarchical decision structure, going from
most general to most specific.  For example, a general rule might
state that, ``If X is a bird, then X can fly'', whereas a more specifc
one might state, ``If X is a bird with small wings and a large body,
then X cannot fly'' \cite{InductionBook}.  When faced with a bird in
the environment, both rules might compete to post their conclusion,
but the more specific one would be favored if both of its conditions are met.

Holland envisioned that, with a complex environment and enough time
for learning, a classifier system might learn an intricate default
hierarchy of rules, and that the competition inherent in the learning
and action mechanisms would allow the system to adapt to a continually
changing environment without losing what it has learned in the past.
Holland put it this way: ``Competition among rules provides the system
with a graceful way of handling perpetual novelty.  When a system has
strong rules that respond to a particular situation, that is the
equivalent of saying that it has certain well-validated hypotheses.
Offspring rules, which begin life weaker than do their parents, can
win the competition and influence the system's behavior only when
there are not strong rules whose conditions are satisfied---in other
words when the system does not know what to do.  If their actions
help, they survive; if not they are soon replaced.  Thus, the
offspring do not interfere with the system's action in well-practiced
situations but wait gracefully in the wings as hypotheses about what
to do under novel circumstances." \cite{HollandSciAmArticle}.

[MM: Default hierarchy emerged in Goldberg's application?  See Holland's discussion.]

@@

begin{itemize}
\item Evolutionary computation
\item Q Learning
\item Backprop
\item    From Turing nomination: "Holland's machine learning system known as the
   Learning Classifier System (LCS), developed in the early 1980s,
   incorporated a reinforcement learning algorithm known as the bucket
   brigade for non-Markovian environments, anticipating by nearly a
   decade non-Markovian learning algorithms."
\end{itemize}
